---
title: "COVID modeling with dgtf"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{COVID modeling with dgtf}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
library(dgtf)
```

# 1. Load Data

```{r}
ytrain <- covid_scz_2020[1:250]
ytest <- covid_scz_2020[251:264]
plot(c(ytrain, ytest), type = "l")
```

# 2. Discretized Hawkes with HVA inference

```{r}
set.seed(123)
mod <- dgtf_model(
  obs = obs_nbinom(),
  link = link_identity(),
  sys = sys_shift(),
  gain = gain_softplus(),
  lag = lag_lognormal(
    meanlog = rnorm(1, 1, 1),
    sigma2 = 1 / rgamma(1, 10, 1)
  ),
  err = err_gaussian(W = 1 / rgamma(1, 10, 1)),
  seasonality = seas_period(7, init = runif(7, 0, 10)),
  zi = FALSE
)

prior <- dgtf_prior(
  seasonality = normal(1, 1),
  W           = inv_gamma(1, 1),
  rho         = inv_gamma(1, 1),
  lag         = list(par1 = normal(0, 1), par2 = inv_gamma(1, 1))
)

fit <- dgtf(
  y       = ytrain,
  model   = mod,
  prior   = prior,
  method  = "hva",
  control = vb_control(iter = 1000, n_sample = 5000, n_particle = 500)
)
attr(fit, "ppc") <- posterior_predict(fit)

plot(fit, what = "Rt")
plot(fit, what = "seas")
plot(fit, what = "W")
plot(fit, what = "marglik") # Check convergence
summary(fit)
```

DGTF model fit (method = HVA, engine = VB)
--------------------------------------------------
Observation : nbinom + link identity
System      : shift  (gain = softplus, lag = lognorm [window = 30])
Seasonality : period 7 (in_state = FALSE)
Data        : n = 250 observations
Posterior   : 5,000 draws
Elapsed     : 468.33 s
  optim     : 370.46 s
  sampling  : 97.84 s

Static parameters
 parameter    mean  median       sd    q025   q975
         W 0.00854 0.00848 0.000922 0.00687 0.0105
   seas[1]     8.5    8.29     1.92     5.4   12.8
   seas[2]     8.6    8.46     1.55    5.95     12
   seas[3]     4.4    4.13     1.68    1.92   8.34
   seas[4]   0.932   0.762     0.69   0.123    2.7
   seas[5]    0.84   0.679    0.646   0.112   2.44
   seas[6]    12.7    12.5     2.35    8.68   17.9
   seas[7]    6.12    5.89     1.71    3.44   10.2
       rho    18.8    18.7     2.47    14.6   24.3
      par1   0.621   0.632    0.273  0.0537   1.12
      par2    3.85    3.38     1.97    1.59   8.92

Latent state psi (median band)
  median range : [-0.511, 0.856]   IQR median : 0.685

Convergence (HVA marglik trace)
  iterations             : 2000
  late window (last 100) : median 8.69, MAD 2.22
  early window           : median 23.9, MAD 20.6
  MAD ratio (late/early) : 0.108  [substantial stabilization]
  tail drift (in MADs)   : 0.08  [flat]
  stabilized             : iter ~ 297  (rolling-MAD < 2x final MAD)
  last large spike       : iter ~ 1860  (|deviation| > 5x final MAD)

Posterior predictive check
<dgtf posterior predictive summary, level = 95%>

Scalar metrics:
 target         metric value nominal_level
      y     chi-square  1.71              
      y           CRPS   9.7              
      y       coverage  0.94          0.95
      y          width  61.5              
      y interval_score  90.3              
     Rt          width 0.372      

```{r}
plot(fit, what = "yhat")
summary(attr(fit, "ppc"))
```

<dgtf posterior predictive summary, level = 95%>

Scalar metrics:
 target         metric value nominal_level
      y     chi-square  1.71              
      y           CRPS   9.7              
      y       coverage  0.94          0.95
      y          width  61.5              
      y interval_score  90.3              
     Rt          width 0.372              

```{r}
fc <- dgtf_forecast_fit(fit, h = 14, ypred_true = ytest)
plot_dgtf_y(fit, fc)
summary(fc)
```

<dgtf forecast summary: h = 14, level = 95%>
Forecast draws : 500,000

Scoring (vs ypred_true):
     metric value nominal_level
   coverage 0.929          0.95
 chi-square 0.399              
       CRPS   3.1  

```{r}
fc
```

<dgtf forecast: h = 14, level = 95%>
Forecast draws : 500,000

Forecast:
 step lower median upper truth hit
    1     8     18    33    22   *
    2     5     13    26    20   *
    3     6     15    28    22   *
    4     6     15    28    15   *
    5     3     11    22    12   *
    6     1      6    15    12   *
    7     1      5    13     7   *
    8     7     16    31    16   *
    9     4     11    24     6   *
   10     5     13    27     6   *
   11     5     14    27     8   *
   12     3     10    21    10   *
   13     1      6    15     7   *
   14     1      5    13    16    

# 3. MCMC inference

```{r}
fit_mcmc <- dgtf(ytrain, mod, prior,
  method = "mcmc",
  control = mcmc_control(
    iter = 5000,
    warmup = 2000,
    thin = 2
  )
)
attr(fit_mcmc, "ppc") <- posterior_predict(fit_mcmc)
summary(fit_mcmc)
```

DGTF model fit (method = MCMC, engine = MCMC)
--------------------------------------------------
Observation : nbinom + link identity
System      : shift  (gain = softplus, lag = lognorm [window = 30])
Seasonality : period 7 (in_state = FALSE)
Data        : n = 250 observations
Posterior   : 5,000 draws
Elapsed     : 3090.71 s

Static parameters
 parameter    mean  median       sd    q025    q975   ess  rhat
         W 0.00804 0.00801 0.000715 0.00675 0.00955 2,402 1.001
   seas[1]    11.3    11.3     2.02    7.42    15.2   767 1.000
   seas[2]    10.5    10.4     1.65    7.56      14   702 1.000
   seas[3]    7.34     7.3     1.75    4.11      11   727 1.000
   seas[4]    3.05    2.94     1.38    0.73    6.03   484 1.001
   seas[5]     2.6    2.47     1.23   0.629    5.22   502 1.000
   seas[6]    15.7    15.6     2.15    11.6      20 1,050 1.000
   seas[7]    9.22    9.21     1.85    5.59    12.9   563 1.001
       rho    17.3    17.2     2.31    13.2    22.2 1,907 1.001
      par1   0.477   0.487    0.293  -0.126    1.03   571 1.000
      par2    3.08    2.77      1.4     1.3    6.47   378 1.002

Latent state psi (median band)
  median range : [-0.747, 0.705]   IQR median : 0.814

HMC sampler
  acceptance rate      : 0.556
  leapfrog step        : 0.07453
  leapfrog steps       : 27
  max |dH|             : 5.96
  99% |dH|             : 4.183
  mean ||grad log pi|| : 23.23

Disturbance MH
  mean acceptance      : 0.688
  range acceptance     : [0.000, 0.774]
  t with accept < 0.1  : 2

Posterior predictive check
<dgtf posterior predictive summary, level = 95%>

Scalar metrics:
 target         metric value nominal_level
      y     chi-square  1.71              
      y           CRPS   9.7              
      y       coverage 0.952          0.95
      y          width  64.3              
      y interval_score  87.8              
     Rt          width 0.402              

```{r}
plot(fit_mcmc, what = "yhat")
summary(attr(fit_mcmc, "ppc"))
```

<dgtf posterior predictive summary, level = 95%>

Scalar metrics:
 target         metric value nominal_level
      y     chi-square  1.71              
      y           CRPS   9.7              
      y       coverage 0.952          0.95
      y          width  64.3              
      y interval_score  87.8              
     Rt          width 0.402              

```{r}
fc_mcmc <- dgtf_forecast_fit(fit_mcmc, h = 14, ypred_true = ytest)
plot_dgtf_y(fit_mcmc, fc_mcmc)
summary(fc_mcmc)
```

<dgtf forecast summary: h = 14, level = 95%>
Forecast draws : 500,000

Scoring (vs ypred_true):
     metric value nominal_level
   coverage     1          0.95
 chi-square 0.304              
       CRPS  3.12    

```{r}
fc_mcmc
```

<dgtf forecast: h = 14, level = 95%>
Forecast draws : 500,000

Forecast:
 step lower median upper truth hit
    1     9     20    37    22   *
    2     6     15    29    20   *
    3     7     17    32    22   *
    4     6     16    32    15   *
    5     4     13    27    12   *
    6     2      8    20    12   *
    7     1      7    18     7   *
    8     8     19    36    16   *
    9     5     14    30     6   *
   10     6     16    33     6   *
   11     6     16    32     8   *
   12     4     12    27    10   *
   13     2      8    20     7   *
   14     1      6    18    16   *

# 4. Compare HVA and MCMC fits

```{r}
dgtf_compare_plot(
  list(HVA = fit, MCMC = fit_mcmc),
  what = "yhat"
)

dgtf_compare_plot(
  list(HVA = fit, MCMC = fit_mcmc),
  what = "Rt"
)

dgtf_compare_plot(
  list(HVA = fit, MCMC = fit_mcmc),
  what = "seas"
)

```

```{r}
dgtf_compare(list(HVA = fit, MCMC = fit_mcmc))
```

<dgtf_comparison: 2 models, level = 95%>
 model n_param log_marglik crps_y chi_y coverage_y width_y interval_score_y elapsed_total elapsed_optim elapsed_sample
   HVA      11        8.69    9.7  1.71       0.94    61.5             90.3        468.3s        370.5s          97.8s
  MCMC      11          NA    9.7  1.71      0.952    64.3             87.8       3090.7s                             

Note: `log_marglik` is the HVA SMC tail estimate of
log p(y | gamma_final); MCMC fits have no equivalent stored
and show NA. Pointwise log-likelihood is not currently
exposed by either engine.

```{r}
dgtf_compare(list(HVA = fc, MCMC = fc_mcmc))
```

<dgtf_comparison (dgtf_forecast): 2 rows, level = 95%>
 model horizon n_draws coverage   chi crps
   HVA      14 500,000    0.929 0.399  3.1
  MCMC      14 500,000        1 0.304 3.12

```{r}
dgtf_compare_params(list(HVA = fit, MCMC = fit_mcmc))
```

<dgtf_param_comparison: 11 parameters x 2 models, level = 95%>
 parameter model  mean median    sd   q_lo  q_hi
      par1   HVA 0.621  0.632 0.273  0.054  1.12
      par1  MCMC 0.477  0.487 0.293 -0.126  1.02
      par2   HVA  3.85   3.38  1.97   1.59  8.92
      par2  MCMC  3.08   2.77   1.4    1.3  6.47
       rho   HVA  18.8   18.7  2.47   14.6  24.3
       rho  MCMC  17.3   17.2  2.31   13.2  22.2
   seas[1]   HVA   8.5   8.29  1.92    5.4  12.8
   seas[1]  MCMC  11.3   11.3  2.02   7.42  15.2
   seas[2]   HVA   8.6   8.46  1.55   5.95    12
   seas[2]  MCMC  10.5   10.4  1.65   7.56    14
   seas[3]   HVA   4.4   4.12  1.68   1.92  8.35
   seas[3]  MCMC  7.34    7.3  1.75   4.11    11
   seas[4]   HVA 0.932  0.762  0.69  0.123  2.71
   seas[4]  MCMC  3.04   2.94  1.38   0.73  6.03
   seas[5]   HVA  0.84  0.679 0.646  0.112  2.44
   seas[5]  MCMC   2.6   2.47  1.23  0.629  5.22
   seas[6]   HVA  12.7   12.5  2.35   8.68  17.9
   seas[6]  MCMC  15.7   15.6  2.15   11.6    20
   seas[7]   HVA  6.12   5.89  1.71   3.44  10.2
   seas[7]  MCMC  9.22   9.21  1.85   5.59  12.9
         W   HVA 0.009  0.008 0.001  0.007 0.011
         W  MCMC 0.008  0.008 0.001  0.007  0.01

# 5. EpiEstim baseline

```{r}
deseason_counts <- function(y, period = 7L, start_day = 1L) {
  n <- length(y)
  day_idx <- ((start_day - 1L + seq_len(n) - 1L) %% period) + 1L
  sds <- vapply(seq_len(period), function(d) {
    sd(y[day_idx == d], na.rm = TRUE)
  }, numeric(1))
  sds <- sds / sum(sds) * period
  round(y / sds[day_idx])
}

y2 <- deseason_counts(ytrain, period = 7L)

epi_baseline <- epiestim_baseline(
  y       = ytrain,
  mean_si = 4.7,
  sd_si   = 2.9,
  method  = "parametric_si",
  window  = 7
)
epi_baseline
```

<epiestim_baseline: method = parametric_si, window = 7, align = center, level = 0.95>
  SI mean = 4.7, SI sd = 2.9
  250 time points (7 NA, 243 estimated)

Last 6 rows:
           lower    median    upper
[245,] 0.6734591 0.8140501 0.972971
[246,] 0.6978092 0.8442140 1.009797
[247,] 0.7415577 0.8955997 1.069632
[248,]        NA        NA       NA
[249,]        NA        NA       NA
[250,]        NA        NA       NA

```{r}
dgtf_compare_plot(
  list(
    HVA = fit,
    MCMC = fit_mcmc,
    EpiEstim = epi_baseline
  ),
  what = "Rt"
)
```


# 6. Compare multiple models

```{r}
set.seed(123)
dl_prior <- dgtf_prior(
  seasonality = normal(0, 10),
  W           = inv_gamma(1, 1),
  rho         = inv_gamma(1, 1),
  lag         = list(par1 = dist_beta(1, 1))
)
vb_opts = vb_control(iter = 1000, n_sample = 5000, n_particle = 500)

dl2 <- dgtf(
  y = ytrain,
  model = dgtf_distributed_lag(
    r = 2,
    kappa = runif(1, 0.1, 0.9),
    seasonality = seas_weekly(init = runif(7, 0, 10))
  ),
  prior = dl_prior,
  method = "hva",
  control = vb_opts
)
attr(dl2, "ppc") <- posterior_predict(dl2, nrep = 100)

summary(dl2)
```

DGTF model fit (method = HVA, engine = VB)
--------------------------------------------------
Observation : nbinom + link identity
System      : shift  (gain = softplus, lag = nbinomp [window = 30])
Seasonality : period 7 (in_state = FALSE)
Data        : n = 250 observations
Posterior   : 5,000 draws
Elapsed     : 160.59 s
  optim     : 106.95 s
  sampling  : 53.62 s

Static parameters
 parameter    mean  median       sd    q025   q975
         W 0.00892 0.00888 0.000825 0.00741 0.0106
   seas[1]    8.92    8.74     1.97    5.64   13.1
   seas[2]    8.92    8.78     1.41    6.56   12.1
   seas[3]    5.67    5.41     1.98    2.64   10.3
   seas[4]    1.38    1.01     1.33  0.0496      5
   seas[5]    1.22   0.619      1.7 0.00417   5.91
   seas[6]    17.2      17     2.49    12.9   22.6
   seas[7]    8.01     7.8     2.04    4.67   12.6
       rho      15    14.9     2.24    11.1   19.8
      par1   0.257   0.224    0.179  0.0143   0.65

Latent state psi (median band)
  median range : [-0.202, 0.622]   IQR median : 0.487

Convergence (HVA marglik trace)
  iterations             : 1000
  late window (last 50 ) : median 11.5, MAD 4.78
  early window           : median 18.9, MAD 11.4
  MAD ratio (late/early) : 0.42  [substantial stabilization]
  tail drift (in MADs)   : -0.07  [flat]
  stabilized             : iter ~ 100  (rolling-MAD < 2x final MAD)
  last large spike       : iter ~ 191  (|deviation| > 5x final MAD)

```{r}
summary(attr(dl2, "ppc"))
plot(dl2, what = "yhat")
```

<dgtf posterior predictive summary, level = 95%>

Scalar metrics:
 target         metric value nominal_level
      y     chi-square  1.76              
      y           CRPS  9.83              
      y       coverage 0.956          0.95
      y          width  74.2              
      y interval_score  89.9              
     Rt          width 0.313              

```{r}
set.seed(123)
ar_prior <- dgtf_prior(
  seasonality = normal(0, 10),
  W = inv_gamma(1, 1),
  rho = inv_gamma(1, 1)
)
ar_init <- dgtf_nb_ar(
  p = 3,
  W = 1 / rgamma(1, 10, 1),
  seasonality = seas_weekly(init = runif(7, 0, 10))
)
ar_fit <- dgtf(
  y       = ytrain,
  model   = ar_init,
  prior   = ar_prior,
  method  = "hva",
  control = vb_control(iter = 1000, n_sample = 5000, n_particle = 500)
)
attr(ar_fit, "ppc") <- posterior_predict(ar_fit, nrep = 100) # truth in a sim study

summary(ar_fit)
```

r$> summary(ar_fit)
DGTF model fit (method = HVA, engine = VB)
--------------------------------------------------
Observation : nbinom + link identity
System      : identity  (gain = softplus, lag = uniform [window = 3])
Seasonality : period 7 (in_state = FALSE)
Data        : n = 250 observations
Posterior   : 5,000 draws
Elapsed     : 386.77 s
  optim     : 257.00 s
  sampling  : 129.76 s

Static parameters
 parameter    mean  median       sd    q025    q975
         W 0.00325 0.00324 0.000211 0.00286 0.00368
   seas[1]    6.57    6.37      1.8    3.71    10.8
   seas[2]    7.68    7.52      1.6    4.99    11.2
   seas[3]    1.19   0.735     1.46  0.0542    4.98
   seas[4]   0.214  0.0846    0.339 0.00054    1.19
   seas[5]   0.411   0.204    0.535 0.00184    1.91
   seas[6]    12.3    12.1     1.77    9.21    16.1
   seas[7]    5.86    5.62     1.82    3.07    10.1
       rho    10.5    10.4     1.41    7.92    13.4

Latent state psi (median band)
  median range : [-1.66, -0.807]   IQR median : 0.426

Convergence (HVA marglik trace)
  iterations             : 1000
  late window (last 50 ) : median 11.1, MAD 3.48
  early window           : median 13.9, MAD 17.4
  MAD ratio (late/early) :  0.2  [substantial stabilization]
  tail drift (in MADs)   : -0.03  [flat]
  stabilized             : iter ~ 170  (rolling-MAD < 2x final MAD)
  last large spike       : iter ~ 333  (|deviation| > 5x final MAD)

```{r}
summary(attr(ar_fit, "ppc"))
plot(ar_fit, what = "yhat")
```

<dgtf posterior predictive summary, level = 95%>

Scalar metrics:
 target         metric value nominal_level
      y     chi-square  2.71              
      y           CRPS  13.4              
      y       coverage 0.916          0.95
      y          width  86.2              
      y interval_score   116              
     Rt          width     0              

```{r}
dgtf_compare_plot(
  list(
    DH = fit,
    DL = dl_fit,
    AR = ar_fit
  ),
  what = "yhat"
)

dgtf_compare_plot(
  list(
    DH = fit,
    DL = dl_fit,
    EpiEstim = epi_baseline
  ),
  what = "Rt"
)

dgtf_compare_plot(
  list(
    DH = fit,
    DL = dl_fit,
    AR = ar_fit
  ),
  what = "seas"
)
```

```{r}
dgtf_compare(
  list(
    DH = fit,
    DL = dl2,
    AR = ar_fit
  )
)
```

<dgtf_comparison: 3 models, level = 95%>
 model n_param log_marglik crps_y chi_y coverage_y width_y interval_score_y elapsed_total elapsed_optim elapsed_sample
    DH      11        8.69    9.7  1.71       0.94    61.5             90.3        468.3s        370.5s          97.8s
    DL      10        11.5   9.83  1.76      0.956    74.2             89.9        160.6s        106.9s          53.6s
    AR       9        11.1   13.4  2.71      0.916    86.2              116        386.8s        257.0s         129.8s

```{r}
set.seed(123)
dh_fc <- dgtf_forecast_fit(fit, h = 14, ypred_true = ytest)
dh_fc
```

<dgtf forecast: h = 14, level = 95%>
Forecast draws : 500,000

Forecast:
 step lower median upper truth hit
    1     8     18    34    22   *
    2     5     13    26    20   *
    3     6     15    28    22   *
    4     6     15    28    15   *
    5     3     11    22    12   *
    6     1      6    15    12   *
    7     1      5    13     7   *
    8     7     16    31    16   *
    9     4     11    24     6   *
   10     5     13    27     6   *
   11     5     14    27     8   *
   12     3     10    21    10   *
   13     1      6    15     7   *
   14     1      5    13    16    

```{r}
summary(dh_fc)
plot_dgtf_y(fit, dh_fc)
```

<dgtf forecast summary: h = 14, level = 95%>
Forecast draws : 500,000

Scoring (vs ypred_true):
     metric value nominal_level
   coverage 0.929          0.95
 chi-square 0.399              
       CRPS   3.1              

```{r}
set.seed(123)
dl_fc <- dgtf_forecast_fit(dl2, h = 14, ypred_true = ytest)
dl_fc
```

<dgtf forecast: h = 14, level = 95%>
Forecast draws : 500,000

Forecast:
 step lower median upper truth hit
    1    10     23    42    22   *
    2     7     18    37    20   *
    3     8     20    40    22   *
    4     8     20    40    15   *
    5     6     17    36    12   *
    6     3     12    29    12   *
    7     2      9    24     7   *
    8     9     23    45    16   *
    9     7     18    38     6    
   10     8     20    41     6    
   11     8     20    41     8   *
   12     6     17    37    10   *
   13     3     12    29     7   *
   14     2      9    24    16   *

```{r}
summary(dl_fc)
plot_dgtf_y(dl2, dl_fc)
```

<dgtf forecast summary: h = 14, level = 95%>
Forecast draws : 500,000

Scoring (vs ypred_true):
     metric value nominal_level
   coverage 0.857          0.95
 chi-square 0.263              
       CRPS  4.17           

```{r}
set.seed(123)
ar_fc <- dgtf_forecast_fit(ar_fit, h = 14, ypred_true = ytest)
ar_fc
```

<dgtf forecast: h = 14, level = 95%>
Forecast draws : 500,000

Forecast:
 step lower median upper truth hit
    1     8     21    41    22   *
    2     6     17    36    20   *
    3     6     18    39    22   *
    4     7     20    45    15   *
    5     4     14    38    12   *
    6     3     12    34    12   *
    7     2     11    33     7   *
    8     7     21    48    16   *
    9     5     17    44     6   *
   10     6     18    46     6   *
   11     7     21    52     8   *
   12     3     15    45    10   *
   13     2     12    42     7   *
   14     2     11    40    16   *

```{r}
summary(ar_fc)
plot_dgtf_y(ar_fit, ar_fc)
```

<dgtf forecast summary: h = 14, level = 95%>
Forecast draws : 500,000

Scoring (vs ypred_true):
     metric value nominal_level
   coverage     1          0.95
 chi-square 0.263              
       CRPS  3.93              

```{r}
dgtf_compare(
  list(
    DH = dh_fc,
    DL = dl_fc,
    AR = ar_fc
  )
)
```

<dgtf_comparison (dgtf_forecast): 3 rows, level = 95%>
 model horizon n_draws coverage   chi crps
    DH      14 500,000    0.929 0.399  3.1
    DL      14 500,000    0.857 0.263 4.17
    AR      14 500,000        1 0.263 3.93