---
title: "Getting started with dgtf"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with dgtf}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6,
                      fig.height = 4)
```

```{r compile-install-package, eval = FALSE}
# # Compile and install this package locally
# Rcpp::compileAttributes()
# devtools::document()
# devtools::install()
```

```{r setup}
library(dgtf)
```

# 1. The DGTF model class
`
A **dynamic generalised transfer function (DGTF)** model has three pieces:

- an *observation* distribution (Poisson, negative-binomial, ...),
- a *link* function and *transfer / lag* structure that connects past
  observations and the latent state to the current intensity, and
- a *system* equation governing how the latent state evolves.

In `dgtf` each piece is its own component object, so you build a model
by snapping the parts together:

```{r build-model}
mod <- dgtf_model(
  obs         = obs_nbinom(),
  link        = link_identity(),
  sys         = sys_shift(),
  gain        = gain_softplus(),
  lag         = lag_lognormal(meanlog = 1.386262, sigma2 = 0.3226017),
  err         = err_gaussian(W = 0.01),
  seasonality = seas_period(1, init = 2),
  zi          = FALSE
)
mod
```

Most components also accept a string shortcut; the call above is
equivalent to `dgtf_model(obs = "nbinom", link = "identity",
sys = "shift", gain = "softplus", lag = lag_lognormal(...), ...)`.

# 2. Simulate

```{r simulate}
sim <- dgtf_simulate_model(mod, ntime = 200, seed = 3269)
print(sim)
# plot(sim, what = "y")
# plot(sim, what = "Rt")
```

# 3. Inference

The "what to infer" decision follows a single rule: anything you give a
prior is treated as unknown; anything you don't is fixed at the value
you supplied to `dgtf_model()`.

```{r prior}
prior <- dgtf_prior(
  seasonality = normal(1, 1),
  W           = inv_gamma(1, 1),
  rho         = inv_gamma(1, 1),
  lag         = list(par1 = normal(0, 1), par2 = inv_gamma(1, 1))
)

```

```{r infer}
fit <- dgtf(
  y       = c(sim$y),
  model   = mod,
  prior   = prior,
  method  = "hva",
  control = vb_control(iter = 1000, n_sample = 5000, n_particle = 500)
)
summary(fit, truth = list(
  W = sim$model$err$params$var, 
  rho = sim$model$obs_init[2],
  par1 = sim$model$lag$params$par1,
  par2 = sim$model$lag$params$par2))
```


# 4. Inspect

```{r}
plot(fit, what = "marglik")
plot(fit, what = "Rt", truth = log1p(exp(sim$psi)))
plot(fit, what = "W")
```

```{r plot, fig.alt = "Posterior intensity vs simulated series"}
# In-sample posterior predictive check
ppc <- posterior_predict(fit, nrep = 100,
                         Rt = log1p(exp(c(sim$psi))))   # truth in a sim study
summary(ppc)
plot(ppc, what = "yhat", truth = c(sim$y), main = "Posterior predictive check")
```

```{r}
fc = dgtf_forecast_fit(fit, h = 5)
print(fc)
```

# 5. Switch the inference method

The `model` and `prior` objects are reusable across methods — only the
control changes:

```{r switch-method, eval = FALSE}
fit_mcmc <- dgtf(sim$y, mod, prior,
                 method  = "mcmc",
                 control = mcmc_control(
                  iter = 5000, 
                  warmup = 2000, 
                  thin = 2, 
                  epsilon = 0.005, 
                  L = 50))

summary(fit_mcmc, truth = list(
  W = sim$model$err$params$var, 
  rho = sim$model$obs_init[2],
  par1 = sim$model$lag$params$par1,
  par2 = sim$model$lag$params$par2))

```

```{r}
plot(fit_mcmc, what = "Rt", truth = log1p(exp(sim$psi)))
plot(fit_mcmc, what = "W", truth = 0.01, type = "trace")
plot(fit_mcmc, what = "W", truth = 0.01, type = "hist")

plot(fit_mcmc, what = "seas", truth = c(1, 2), type = "trace")

plot(fit_mcmc, what = "par1", truth = 1.386262, type = "trace")
plot(fit_mcmc, what = "par2", truth = 0.3226017, type = "hist")
```

```{r plot, fig.alt = "Posterior intensity vs simulated series"}
# In-sample posterior predictive check
ppc_mcmc <- posterior_predict(fit_mcmc, nrep = 100,
                         Rt = log1p(exp(c(sim$psi))))   # truth in a sim study
summary(ppc_mcmc)
plot(ppc_mcmc, what = "yhat", truth = c(sim$y), main = "Posterior predictive check")
```

```{r}
dgtf_compare_plot(
    list(HVA = ppc, MCMC = ppc_mcmc),
    what = "yhat"
)

dgtf_compare_plot(
    list(HVA = fit, MCMC = fit_mcmc,
         True = log1p(exp(sim$psi))),
    what = "Rt"
)

dgtf_compare_plot(
    list(HVA = fit, MCMC = fit_mcmc,
         True = 2),
    what = "seas"
)

dgtf_compare_plot(
    list(HVA = fit, MCMC = fit_mcmc,
         True = 0.01),
    what = "W"
)

dgtf_compare_plot(
    list(HVA = fit, MCMC = fit_mcmc,
         True = 1.386262),
    what = "par1"
)

dgtf_compare_plot(
    list(HVA = fit, MCMC = fit_mcmc,
         True = 0.3226017),
    what = "par2"
)

dgtf_compare_plot(
    list(HVA = fit, MCMC = fit_mcmc,
         True = 30),
    what = "rho"
)
```

# 6. Model comparison

For the canonical models discussed in the manuscript there are
one-line shortcuts:

```{r canonical}
m0_init <- dgtf_hawkes(
  meanlog = runif(1, 1, 2),
  sigma2 = runif(1, 0.1, 0.5),
  W = 1 / rgamma(1, 10, 1),
  seasonality = seas_period(1, init = runif(1, 0, 10))
)
m1_init = m0_init


m2_prior <- dgtf_prior(
  seasonality = normal(1, 1),
  W           = inv_gamma(1, 1),
  rho         = inv_gamma(1, 1),
  lag         = list(par1 = dist_beta(1.5, 2))
)

m2_init = dgtf_distributed_lag(
  r = 4, 
  kappa = 0.395,
  seasonality = seas_period(2, init = c(1, 2))
)
m2_fit <- dgtf(
  y       = c(sim$y),
  model   = m2_init,
  prior   = m2_prior,
  method  = "hva",
  control = vb_control(iter = 1000, n_sample = 5000, n_particle = 500)
)
m2_ppc <- posterior_predict(m2_fit, nrep = 100,
                         Rt = log1p(exp(c(sim$psi))))   # truth in a sim study
summary(m2_ppc)


m3_prior <- dgtf_prior(
  seasonality = normal(1, 1),
  W = inv_gamma(1, 1),
  rho = inv_gamma(1, 1)
)
m3_init = dgtf_nb_ar(
  p = 3,
  W = 1 / rgamma(1, 10, 1),
  seasonality = seas_period(1, init = runif(1, 0, 10))
)
m3_fit <- dgtf(
  y       = c(sim$y),
  model   = m3_init,
  prior   = m3_prior,
  method  = "hva",
  control = vb_control(iter = 1000, n_sample = 5000, n_particle = 500)
)
m3_ppc <- posterior_predict(m3_fit, nrep = 100,
                         Rt = log1p(exp(c(sim$psi))))   # truth in a sim study
summary(m3_ppc)


m4_prior <- dgtf_prior(
  seasonality = normal(1, 1),
  W = inv_gamma(1, 1)
)
m4_init  = dgtf_poisson_ar(
  p = 5,
  W = 1 / rgamma(1, 10, 1),
  seasonality = seas_period(1, init = runif(1, 0, 10))
)
m4_fit <- dgtf(
  y       = c(sim$y),
  model   = m4_init,
  prior   = m4_prior,
  method  = "hva",
  control = vb_control(iter = 1000, n_sample = 5000, n_particle = 500)
)
```


# 7. Swap a component to change the model

The same data can be analysed under a different DGTF variant by
substituting one component:

```{r swap, eval = FALSE}
mod_geo  <- dgtf_hawkes(lag = lag_nbinom(r = 1, kappa = 0.5, window = 30))
mod_relu <- dgtf_hawkes(gain = gain_ramp())
```

A side-by-side comparison of all four canonical models on the same
dataset will appear in `vignette("four-models")`.
