---
title: "Getting started with dgtf"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with dgtf}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 6,
  fig.height = 4
)
```

```{r compile-install-package, eval = FALSE}
# # Compile and install this package locally
# Rcpp::compileAttributes()
# devtools::document()
# devtools::install()
```

```{r setup}
library(dgtf)
library(ggplot2)
```

# 1. The DGTF model class
`
A **dynamic generalised transfer function (DGTF)** model has three pieces:

- an *observation* distribution (Poisson, negative-binomial, ...),
- a *link* function and *transfer / lag* structure that connects past
  observations and the latent state to the current intensity, and
- a *system* equation governing how the latent state evolves.

In `dgtf` each piece is its own component object, so you build a model
by snapping the parts together:

```{r build-model}
mod <- dgtf_model(
  obs         = obs_nbinom(),
  link        = link_identity(),
  sys         = sys_shift(),
  gain        = gain_softplus(),
  lag         = lag_lognormal(meanlog = 1.35, sigma2 = 0.32),
  err         = err_gaussian(W = 0.01),
  seasonality = seas_period(1, init = 2),
  zi          = FALSE
)
mod
```

Most components also accept a string shortcut; the call above is
equivalent to `dgtf_model(obs = "nbinom", link = "identity",
sys = "shift", gain = "softplus", lag = lag_lognormal(...), ...)`.

# 2. Simulate

```{r simulate}
sim <- dgtf_simulate_model(mod, ntime = 210, seed = 3269)
ytrain <- c(sim$y)[1:200]
ytest <- c(sim$y)[201:210]
Rt_train <- log1p(exp(c(sim$psi)))[1:200]
Rt_test <- log1p(exp(c(sim$psi)))[201:210]

print(sim)
plot(sim, what = "y")
plot(sim, what = "Rt")
```

# 3. Inference

The "what to infer" decision follows a single rule: anything you give a
prior is treated as unknown; anything you don't is fixed at the value
you supplied to `dgtf_model()`.

```{r prior}
prior <- dgtf_prior(
  seasonality = normal(1, 1),
  W           = inv_gamma(1, 1),
  rho         = inv_gamma(1, 1),
  lag         = list(par1 = normal(0, 1), par2 = inv_gamma(1, 1))
)

```

```{r infer}
set.seed(123)
m0_init <- dgtf_hawkes(
  meanlog = runif(1, 1, 2),
  sigma2 = runif(1, 0.1, 0.5),
  W = 1 / rgamma(1, 10, 1),
  seasonality = seas_period(1, init = runif(1, 1, 10))
)


fit <- dgtf(
  y       = ytrain,
  model   = m0_init,
  prior   = prior,
  method  = "hva",
  control = vb_control(iter = 1000, n_sample = 5000, n_particle = 500)
)
summary(fit, truth = list(
  W = sim$model$err$params$var,
  seas = sim$model$seasonality$init,
  rho = sim$model$obs_init[2],
  par1 = sim$model$lag$params$par1,
  par2 = sim$model$lag$params$par2
))
```


# 4. Inspect

```{r}
plot(fit, what = "marglik")
plot(fit, what = "Rt", truth = Rt_train)
plot(fit, what = "W")
```

```{r plot, fig.alt = "Posterior intensity vs simulated series"}
# In-sample posterior predictive check
ppc <- posterior_predict(fit, nrep = 100, Rt = Rt_train)   # truth in a sim study
attr(fit, "ppc") <- ppc
summary(ppc)
p = plot(ppc, what = "yhat", truth = ytrain)
```

```{r}
fc = dgtf_forecast_fit(fit, h = 10, ypred_true = ytest)
summary(fc)
plot_dgtf_y(fit, fc)
# plot_dgtf_y(ppc, fc, truth_in = fit$y)
plot(fc)
```

# 5. Switch the inference method

The `model` and `prior` objects are reusable across methods — only the
control changes:

```{r switch-method, eval = FALSE}
set.seed(123)

fit_mcmc <- dgtf(ytrain, mod, prior,
  method = "mcmc",
  control = mcmc_control(
    iter = 5000,
    warmup = 2000,
    thin = 5,
    epsilon = 0.005,
    L = 50
  )
)

summary(fit_mcmc, truth = list(
  W = sim$model$err$params$var,
  seas = sim$model$seasonality$init,
  rho = sim$model$obs_init[2],
  par1 = sim$model$lag$params$par1,
  par2 = sim$model$lag$params$par2
))

```

```{r}
plot(fit_mcmc, what = "Rt", truth = log1p(exp(sim$psi)))
plot(fit_mcmc, what = "W", truth = 0.01, type = "trace")
plot(fit_mcmc, what = "W", truth = 0.01, type = "hist")

plot(fit_mcmc, what = "seas", truth = c(1, 2), type = "trace")

plot(fit_mcmc, what = "par1", truth = 1.386262, type = "trace")
plot(fit_mcmc, what = "par2", truth = 0.3226017, type = "hist")
```

```{r plot, fig.alt = "Posterior intensity vs simulated series"}
# In-sample posterior predictive check
ppc_mcmc <- posterior_predict(fit_mcmc, nrep = 100, Rt = Rt_train)   # truth in a sim study
summary(ppc_mcmc)
plot(ppc_mcmc, what = "yhat", truth = ytrain, main = "Posterior predictive check")

fc_mcmc = dgtf_forecast_fit(fit_mcmc, h = 10, ypred_true = ytest)
print(fc_mcmc)
```

```{r}
dgtf_compare_plot(
    list(HVA = ppc, MCMC = ppc_mcmc),
    what = "yhat"
)

dgtf_compare_plot(
    list(HVA = fit, MCMC = fit_mcmc,
         True = log1p(exp(sim$psi))),
    what = "Rt"
)

dgtf_compare_plot(
    list(HVA = fit, MCMC = fit_mcmc,
         True = 2),
    what = "seas"
)

dgtf_compare_plot(
    list(HVA = fit, MCMC = fit_mcmc,
         True = 0.01),
    what = "W"
)

dgtf_compare_plot(
    list(HVA = fit, MCMC = fit_mcmc,
         True = 1.386262),
    what = "par1"
)

dgtf_compare_plot(
    list(HVA = fit, MCMC = fit_mcmc,
         True = 0.3226017),
    what = "par2"
)

dgtf_compare_plot(
    list(HVA = fit, MCMC = fit_mcmc,
         True = 30),
    what = "rho"
)
```

```{r}
truth_list <- list(
    W    = sim$model$err$params$var,
    seas = sim$model$seasonality$init,
    rho  = sim$model$obs_init[2],
    par1 = sim$model$lag$params$par1,
    par2 = sim$model$lag$params$par2
)

dgtf_compare(list(HVA = fit, MCMC = fit_mcmc), truth = truth_list)
dgtf_compare_params(list(HVA = fit, MCMC = fit_mcmc), truth = truth_list)
```


# 6. Model comparison

For the canonical models discussed in the manuscript there are
one-line shortcuts:

```{r canonical}
set.seed(123)
m1_fit <- dgtf(
  y = ytrain,
  model = mod,
  prior = dgtf_prior(
    seasonality = normal(1, 1),
    rho = inv_gamma(1, 1),
    lag = list(
      par1 = normal(0, 1),
      par2 = inv_gamma(1, 1)
    )
  ),
  method = "hva",
  control = vb_control(
    iter = 1000,
    n_sample = 5000,
    n_particle = 500,
    use_discount = TRUE,
    discount_factor = 0.92
  )
)
plot(m1_fit, what = "Rt")

m1_ppc <- posterior_predict(m1_fit, nrep = 100, Rt = Rt_train) # truth in a sim study
summary(m1_ppc)
m1_fc <- dgtf_forecast_fit(m1_fit, h = 3, ypred_true = ytest[1:3])

m2_prior <- dgtf_prior(
  seasonality = normal(1, 1),
  W           = inv_gamma(1, 1),
  rho         = inv_gamma(1, 1),
  lag         = list(par1 = dist_beta(1.5, 2))
)

m2_init <- dgtf_distributed_lag(
  r = 4,
  kappa = 0.395,
  seasonality = seas_period(2, init = c(1, 2))
)
m2_fit <- dgtf(
  y       = ytrain,
  model   = m2_init,
  prior   = m2_prior,
  method  = "hva",
  control = vb_control(iter = 1000, n_sample = 5000, n_particle = 500)
)
m2_ppc <- posterior_predict(m2_fit, nrep = 100, Rt = Rt_train) # truth in a sim study
summary(m2_ppc)
plot(m2_ppc, what = "yhat")

set.seed(123)
m3_prior <- dgtf_prior(
  seasonality = normal(1, 1),
  W = inv_gamma(1, 1),
  rho = inv_gamma(1, 1)
)
m3_init <- dgtf_nb_ar(
  p = 3,
  W = 1 / rgamma(1, 10, 1),
  seasonality = seas_period(1, init = runif(1, 0, 10))
)
m3_fit <- dgtf(
  y       = ytrain,
  model   = m3_init,
  prior   = m3_prior,
  method  = "hva",
  control = vb_control(iter = 1000, n_sample = 5000, n_particle = 500)
)
m3_ppc <- posterior_predict(m3_fit, nrep = 100, Rt = Rt_train) # truth in a sim study
summary(m3_ppc)
plot(m3_ppc, what = "yhat")


set.seed(123)
m4_prior <- dgtf_prior(
  seasonality = normal(1, 1),
  W = inv_gamma(1, 1)
)
m4_init <- dgtf_poisson_ar(
  p = 5,
  W = 0.001,
  seasonality = seas_period(1, init = runif(1, 0, 10))
)
m4_fit <- dgtf(
  y       = ytrain,
  model   = m4_init,
  prior   = m4_prior,
  method  = "hva",
  control = vb_control(iter = 1000, n_sample = 5000, n_particle = 500)
)
m4_ppc <- posterior_predict(m4_fit,
  nrep = 100,
  Rt = log1p(exp(c(sim$psi)))
) # truth in a sim study
summary(m4_ppc)
plot(m4_ppc, what = "yhat")

```

```{r}
dgtf_compare_plot(
  list(
    M0 = ppc,
    M1 = m1_ppc,
    M2 = m2_ppc,
    M3 = m3_ppc,
    M4 = m4_ppc,
    True = c(sim$y)
  ),
  what = "yhat"
)

dgtf_compare_plot(
  list(
    M0 = fit,
    M1 = m1_fit,
    M2 = m2_fit,
    True = Rt_train
  ),
  what = "Rt"
)

dgtf_compare(list(
  M0 = fit,
  M1 = m1_fit,
  M2 = m2_fit,
  M3 = m3_fit,
  M4 = m4_fit
))
```

# 7. EpiEstim baseline

For comparison with a non-state-space estimator, the package's `epiestim_baseline()` wrapper runs [`EpiEstim::estimate_R()`][EpiEstim] with a sliding window and returns the `R_t` posterior bands in a format that `dgtf_compare_plot()` consumes directly. `EpiEstim` is in `Suggests`, so this section is conditional on having the package installed.

[EpiEstim]: https://cran.r-project.org/package=EpiEstim

```{r epiestim, eval = requireNamespace("EpiEstim", quietly = TRUE)}
# Serial-interval mean / sd implied by the Hawkes simulator's
# lognormal kernel `dgtf_hawkes(meanlog = 1.386, sigma2 = 0.323)`:
#   mean = exp(meanlog + sigma2 / 2)      ~= 4.7
#   sd   = mean * sqrt(exp(sigma2) - 1)   ~= 2.9
epi_baseline <- epiestim_baseline(
  y       = ytrain,
  mean_si = 4.7,
  sd_si   = 2.9,
  method  = "parametric_si",
  window  = 7
)
epi_baseline

dgtf_compare_plot(
  list(
    M0       = fit,
    EpiEstim = epi_baseline,
    True     = Rt_train
  ),
  what = "Rt"
)
```