---
title: "Getting started with dgtf"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with dgtf}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 6,
  fig.height = 4
)
```

```{r compile-install-package, eval = FALSE}
# # Compile and install this package locally
# Rcpp::compileAttributes()
# devtools::document()
# devtools::install()
```

```{r setup}
library(dgtf)
library(ggplot2)
```

# 1. The DGTF model class
`
A **dynamic generalised transfer function (DGTF)** model has three pieces:

- an *observation* distribution (Poisson, negative-binomial, ...),
- a *link* function and *transfer / lag* structure that connects past
  observations and the latent state to the current intensity, and
- a *system* equation governing how the latent state evolves.

In `dgtf` each piece is its own component object, so you build a model
by snapping the parts together:

```{r build-model}
mod <- dgtf_model(
  obs         = obs_nbinom(),
  link        = link_identity(),
  sys         = sys_shift(),
  gain        = gain_softplus(),
  lag         = lag_lognormal(meanlog = 1.35, sigma2 = 0.32),
  err         = err_gaussian(W = 0.01),
  seasonality = seas_period(1, init = 2),
  zi          = FALSE
)
mod
```

Most components also accept a string shortcut; the call above is
equivalent to `dgtf_model(obs = "nbinom", link = "identity",
sys = "shift", gain = "softplus", lag = lag_lognormal(...), ...)`.

# 2. Simulate

```{r simulate}
sim <- dgtf_simulate_model(mod, ntime = 210, seed = 3269)
ytrain <- c(sim$y)[1:200]
ytest <- c(sim$y)[201:210]
Rt_train <- log1p(exp(c(sim$psi)))[1:200]
Rt_test <- log1p(exp(c(sim$psi)))[201:210]

print(sim)
plot(sim, what = "y")
plot(sim, what = "Rt")
```

# 3. Inference

The "what to infer" decision follows a single rule: anything you give a
prior is treated as unknown; anything you don't is fixed at the value
you supplied to `dgtf_model()`.

```{r prior}
prior <- dgtf_prior(
  seasonality = normal(0, 10),
  W           = inv_gamma(1, 1),
  rho         = inv_gamma(1, 1),
  lag         = list(par1 = normal(0, 1), par2 = inv_gamma(1, 1))
)

```

```{r infer}
set.seed(123)
m0_init <- dgtf_hawkes(
  meanlog = runif(1, 1, 2),
  sigma2 = runif(1, 0.1, 0.5),
  W = 1 / rgamma(1, 10, 1),
  seasonality = seas_period(1, init = runif(1, 1, 10))
)


fit <- dgtf(
  y       = ytrain,
  model   = m0_init,
  prior   = prior,
  method  = "hva",
  control = vb_control(iter = 1000, n_sample = 5000, n_particle = 500)
)
summary(fit)
```

DGTF model fit (method = HVA)
--------------------------------------------------
Observation : nbinom + link identity
System      : shift  (gain = softplus, lag = lognorm [window = 30])
Seasonality : period 1 (in_state = FALSE)
Data        : n = 200 observations
Posterior   : 5,000 draws
Elapsed     : 152.88 s
  optim     : 102.83 s
  sampling  : 50.04 s

Static parameters
 parameter  mean median      sd    q025   q975
         W 0.012  0.012 0.00139 0.00945 0.0149
      seas  2.28   2.23   0.535     1.4    3.5
       rho  34.6   34.2    4.95    25.7   45.4
      par1  1.06   1.06   0.161   0.735   1.37
      par2 0.389  0.362   0.157   0.177  0.767

Latent state psi (median band)
  median range : [-0.129, 1.25]   IQR median : 0.643

Convergence (HVA marglik trace)
  iterations             : 1000
  late window (last 50 ) : median -0.294, MAD 0.449
  early window           : median 1.54, MAD 3.22
  MAD ratio (late/early) : 0.14  [substantial stabilization]
  tail drift (in MADs)   : 0.27  [flat]
  stabilized             : iter ~ 260  (rolling-MAD < 2x final MAD)
  last large spike       : iter ~ 298  (|deviation| > 5x final MAD)

```{r}
truth_list <- list(W = 0.01, seas = 2, rho = 30, par1 = 1.35, par2 = 0.32)
param_recovery(fit, truth_list, digits = 3)
```

 parameter truth   mean median    sd   q_lo   q_hi   bias   mae  rmse  crps  
         W  0.01  0.012  0.012 0.001  0.009  0.015  0.002 0.002 0.002 0.001 *
      seas  2.00  2.278  2.225 0.535  1.401  3.499  0.278 0.457 0.603 0.160 *
       rho 30.00 34.561 34.190 4.954 25.739 45.368  4.561 5.346 6.733 2.571 *
      par1  1.35  1.057  1.058 0.161  0.735  1.368 -0.293 0.297 0.334 0.206 *
      par2  0.32  0.389  0.362 0.157  0.177  0.767  0.069 0.119 0.171 0.037 *
---
'*' indicates the truth lies within the 95% credible interval

# 4. Inspect

```{r}
plot(fit, what = "marglik")
plot(fit, what = "Rt", truth = Rt_train)
plot(fit, what = "W")

ggsave(file.path(opath,'covid_model_comparison_yhat_yforecast.pdf'), plot = last_plot(), device = "pdf", dpi = 300, width = 7, height = 2)
```

```{r plot, fig.alt = "Posterior intensity vs simulated series"}
# In-sample posterior predictive check
attr(fit, "ppc") <- posterior_predict(fit, nrep = 100, Rt = Rt_train)   # truth in a sim study
plot(fit, what = "yhat", truth = ytrain)
summary(attr(fit, "ppc"))
```

<dgtf posterior predictive summary, level = 95%>

Scalar metrics:
 target         metric value nominal_level
      y     chi-square 0.966              
      y           CRPS  32.5              
      y       coverage 0.965          0.95
      y          width   239              
      y interval_score   249              
     Rt            MAE 0.116              
     Rt           RMSE  0.15              
     Rt       coverage 0.899          0.95
     Rt          width 0.375              
     Rt interval_score 0.486              

```{r}
fc <- dgtf_forecast_fit(fit, h = 10, ypred_true = ytest)
plot_dgtf_y(fit, fc)
plot(fc)
summary(fc)
```

<dgtf forecast summary: h = 10, level = 95%>
Forecast draws : 500,000

Scoring (vs ypred_true):
     metric  value nominal_level
   coverage      1          0.95
 chi-square 0.0964              
       CRPS     24              


# 5. Switch the inference method

The `model` and `prior` objects are reusable across methods — only the
control changes:

```{r switch-method, eval = FALSE}
set.seed(123)

fit_mcmc <- dgtf(ytrain, mod, prior,
  method = "mcmc",
  control = mcmc_control(
    iter = 5000,
    warmup = 2000,
    thin = 5
  )
)

summary(fit_mcmc)
```


DGTF model fit (method = MCMC)
--------------------------------------------------
Observation : nbinom + link identity
System      : shift  (gain = softplus, lag = lognorm [window = 30])
Seasonality : period 1 (in_state = FALSE)
Data        : n = 200 observations
Posterior   : 5,000 draws
Elapsed     : 2821.63 s

Static parameters
 parameter  mean  median      sd    q025   q975   ess  rhat
         W  0.01 0.00992 0.00102 0.00827 0.0122 2,967 1.001
      seas  2.46    2.43   0.525    1.51   3.57   460 1.000
       rho  29.6    29.4    4.52    21.6     39 2,739 1.001
      par1   1.1     1.1    0.17   0.783   1.44   769 1.000
      par2 0.491   0.453   0.201   0.215  0.994   922 1.000

Latent state psi (median band)
  median range : [-0.182, 1.28]   IQR median : 0.706

HMC sampler
  acceptance rate      : 0.732
  leapfrog step        : 0.1845
  leapfrog steps       : 11
  max |dH|             : 3.362e+11
  99% |dH|             : 3.527
  mean ||grad log pi|| : 21.18

Disturbance MH
  mean acceptance      : 0.763
  range acceptance     : [0.000, 0.802]
  t with accept < 0.1  : 2

```{r}
plot(fit_mcmc, what = "Rt", truth = log1p(exp(sim$psi)))
plot(fit_mcmc, what = "W", truth = 0.01, type = "trace")
plot(fit_mcmc, what = "W", truth = 0.01, type = "hist")

plot(fit_mcmc, what = "seas", truth = c(1, 2), type = "trace")

plot(fit_mcmc, what = "par1", truth = 1.386262, type = "trace")
plot(fit_mcmc, what = "par2", truth = 0.3226017, type = "hist")
```

```{r plot, fig.alt = "Posterior intensity vs simulated series"}
# In-sample posterior predictive check
attr(fit_mcmc, "ppc") <- posterior_predict(fit_mcmc, nrep = 100, Rt = Rt_train)   # truth in a sim study
plot(fit_mcmc, what = "yhat", truth = ytrain)
summary(attr(fit_mcmc, "ppc"))
```

<dgtf posterior predictive summary, level = 95%>

Scalar metrics:
 target         metric value nominal_level
      y     chi-square  1.03              
      y           CRPS    33              
      y       coverage 0.975          0.95
      y          width   261              
      y interval_score   263              
     Rt            MAE 0.124              
     Rt           RMSE 0.159              
     Rt       coverage 0.975          0.95
     Rt          width 0.437              
     Rt interval_score 0.459              

```{r}
fc_mcmc = dgtf_forecast_fit(fit_mcmc, h = 10, ypred_true = ytest)
print(fc_mcmc)
```

<dgtf forecast: h = 10, level = 95%>
Forecast draws : 500,000

Forecast:
 step lower median upper truth hit
    1   110    179   276   224   *
    2    95    157   247   178   *
    3    82    142   235   196   *
    4    71    129   227   157   *
    5    61    117   221   154   *
    6    52    106   216   161   *
    7    44     96   211   130   *
    8    37     88   209   133   *
    9    31     80   207    92   *
   10    26     72   206   122   *

```{r}
dgtf_compare_plot(
    list(HVA = fit, MCMC = fit_mcmc),
    what = "yhat"
)

dgtf_compare_plot(
    list(HVA = fit, MCMC = fit_mcmc,
         True = Rt_train),
    what = "Rt"
)

dgtf_compare_plot(
    list(HVA = fit, MCMC = fit_mcmc,
         True = 2),
    what = "seas"
)

dgtf_compare_plot(
    list(HVA = fit, MCMC = fit_mcmc,
         True = 0.01),
    what = "W"
)

dgtf_compare_plot(
    list(HVA = fit, MCMC = fit_mcmc,
         True = 1.35),
    what = "par1"
)

dgtf_compare_plot(
    list(HVA = fit, MCMC = fit_mcmc,
         True = 0.32),
    what = "par2"
)

dgtf_compare_plot(
    list(HVA = fit, MCMC = fit_mcmc,
         True = 30),
    what = "rho"
)
```

```{r}
truth_list <- list(
    W    = sim$model$err$params$var,
    seas = sim$model$seasonality$init,
    rho  = sim$model$obs_init[2],
    par1 = sim$model$lag$params$par1,
    par2 = sim$model$lag$params$par2
)

dgtf_compare(list(HVA = fit, MCMC = fit_mcmc), truth = truth_list)
```

<dgtf_comparison (dgtf_fit): 2 rows, level = 95%>
 model n_param log_marglik crps_y chi_y coverage_y width_y interval_score_y elapsed_total elapsed_optim elapsed_sample n_param_truth param_coverage_rate
   HVA       5      -0.294   32.5 0.966      0.965     239              249        152.9s        102.8s          50.0s             5                   1
  MCMC       5          NA     33  1.03      0.975     261              263       2821.6s                                          5                   1

Note: `log_marglik` is the HVA SMC tail estimate of
log p(y | gamma_final); MCMC fits have no equivalent stored
and show NA. Pointwise log-likelihood is not currently
exposed by either inference method.

Use `dgtf_compare_params(fits, truth)` for the
per-parameter recovery table.

```{r}
dgtf_compare_params(list(HVA = fit, MCMC = fit_mcmc), truth = truth_list)
```

<dgtf_param_comparison: 5 parameters x 2 models, level = 95%>
 parameter model truth  mean median    sd  q_lo  q_hi   bias   mae  rmse  crps  
      par1   HVA  1.35  1.06   1.06 0.161 0.735  1.37 -0.293 0.297 0.334 0.206 *
      par1  MCMC  1.35   1.1    1.1  0.17 0.783  1.44 -0.249 0.262 0.301 0.166 *
      par2   HVA  0.32 0.389  0.362 0.157 0.177 0.767  0.069 0.119 0.171 0.037 *
      par2  MCMC  0.32 0.491  0.453 0.201 0.215 0.994  0.171 0.192 0.264 0.085 *
       rho   HVA    30  34.6   34.2  4.95  25.7  45.4   4.56  5.35  6.73  2.57 *
       rho  MCMC    30  29.6   29.4  4.52  21.6    39 -0.393  3.65  4.53   1.1 *
      seas   HVA     2  2.28   2.23 0.535   1.4   3.5  0.278 0.457 0.603  0.16 *
      seas  MCMC     2  2.46   2.43 0.525  1.51  3.57  0.459 0.555 0.697  0.26 *
         W   HVA  0.01 0.012  0.012 0.001 0.009 0.015  0.002 0.002 0.002 0.001 *
         W  MCMC  0.01  0.01   0.01 0.001 0.008 0.012      0 0.001 0.001     0 *
---
'*' indicates the truth lies within the 95% credible interval

```{r}
param_recovery(fit, truth_list)
```

 parameter truth   mean median    sd   q_lo   q_hi   bias   mae  rmse  crps  
         W  0.01  0.012  0.012 0.001  0.009  0.015  0.002 0.002 0.002 0.001 *
      seas  2.00  2.278  2.225 0.535  1.401  3.499  0.278 0.457 0.603 0.160 *
       rho 30.00 34.561 34.190 4.954 25.739 45.368  4.561 5.346 6.733 2.571 *
      par1  1.35  1.057  1.058 0.161  0.735  1.368 -0.293 0.297 0.334 0.206 *
      par2  0.32  0.389  0.362 0.157  0.177  0.767  0.069 0.119 0.171 0.037 *
---
'*' indicates the truth lies within the 95% credible interval

```{r}
dgtf_compare(list(HVA = fc, MCMC = fc_mcmc))
```

<dgtf_comparison (dgtf_forecast): 2 rows, level = 95%>
 model horizon n_draws coverage    chi crps
   HVA      10 500,000        1 0.0964   24
  MCMC      10 500,000        1  0.123   23

# 6. Model comparison

For the canonical models discussed in the manuscript there are
one-line shortcuts:

```{r canonical}
set.seed(123)
# m1_init <- dgtf_hawkes(
#   meanlog = 1.35,
#   sigma2 = 0.32,
#   seasonality = seas_period(1, init = 2)
# )

m1_fit <- dgtf(
  y = ytrain,
  model = mod,
  prior = dgtf_prior(
    seasonality = normal(0, 10),
    rho = inv_gamma(1, 1),
    lag = list(
      par1 = normal(0, 1),
      par2 = inv_gamma(1, 1)
    )
  ),
  method = "hva",
  control = vb_control(
    iter = 1000,
    n_sample = 5000,
    n_particle = 500,
    use_discount = TRUE,
    discount_factor = 0.92
  )
)
attr(m1_fit, "ppc") <- posterior_predict(m1_fit, nrep = 100, Rt = Rt_train) # truth in a sim study
plot(m1_fit, what = "Rt", truth = Rt_train)
m1_fc <- dgtf_forecast_fit(m1_fit, h = 10, ypred_true = ytest)


m2_prior <- dgtf_prior(
  seasonality = normal(0, 10),
  W           = inv_gamma(1, 1),
  rho         = inv_gamma(1, 1),
  lag         = list(par1 = dist_beta(1, 1))
)
m2_init <- dgtf_distributed_lag(
  r = 4,
  kappa = 0.395,
  seasonality = seas_period(1, init = runif(1, 0, 10))
)
m2_fit <- dgtf(
  y       = ytrain,
  model   = m2_init,
  prior   = m2_prior,
  method  = "hva",
  control = vb_control(iter = 1000, n_sample = 5000, n_particle = 500)
)
attr(m2_fit, "ppc") <- posterior_predict(m2_fit, nrep = 100, Rt = Rt_train) # truth in a sim study
m2_fc <- dgtf_forecast_fit(m2_fit, h = 10, ypred_true = ytest)


set.seed(123)
m3_prior <- dgtf_prior(
  seasonality = normal(0, 10),
  W = inv_gamma(1, 1),
  rho = inv_gamma(1, 1)
)
m3_init <- dgtf_nb_ar(
  p = 3,
  W = 1 / rgamma(1, 10, 1),
  seasonality = seas_period(1, init = runif(1, 0, 10))
)
m3_fit <- dgtf(
  y       = ytrain,
  model   = m3_init,
  prior   = m3_prior,
  method  = "hva",
  control = vb_control(iter = 1000, n_sample = 5000, n_particle = 500)
)
attr(m3_fit, "ppc") <- posterior_predict(m3_fit, nrep = 100, Rt = Rt_train) # truth in a sim study
m3_fc <- dgtf_forecast_fit(m3_fit, h = 10, ypred_true = ytest)


set.seed(123)
m4_prior <- dgtf_prior(
  seasonality = normal(0, 10),
  W = inv_gamma(1, 1)
)
m4_init <- dgtf_poisson_ar(
  p = 5,
  W = 0.001,
  seasonality = seas_period(1, init = runif(1, 0, 10))
)
m4_fit <- dgtf(
  y       = ytrain,
  model   = m4_init,
  prior   = m4_prior,
  method  = "hva",
  control = vb_control(iter = 1000, n_sample = 5000, n_particle = 500)
)
attr(m4_fit, "ppc") <- posterior_predict(m4_fit,
  nrep = 100,
  Rt = Rt_train
) # truth in a sim study
m4_fc <- dgtf_forecast_fit(m4_fit, h = 10, ypred_true = ytest)
```

```{r}
plot_dgtf_y(fit, fc)
plot_dgtf_y(m1_fit, m1_fc)

dgtf_compare_plot(
  list(
    M0 = fit,
    M1 = m1_fit,
    M2 = m2_fit,
    M3 = m3_fit,
    M4 = m4_fit,
    True = ytrain
  ),
  what = "yhat"
)

dgtf_compare_plot(
  list(
    M0 = fit,
    M1 = m1_fit,
    M2 = m2_fit,
    True = Rt_train
  ),
  what = "Rt"
)
```

```{r}
dgtf_compare(list(
  M0 = fit,
  M1 = m1_fit,
  M2 = m2_fit,
  M3 = m3_fit,
  M4 = m4_fit
), Rt_truth = Rt_train)
```

<dgtf_comparison (dgtf_fit): 5 rows, level = 95%>
 model n_param log_marglik crps_y chi_y coverage_y width_y interval_score_y elapsed_total elapsed_optim elapsed_sample mae_Rt rmse_Rt coverage_Rt interval_score_Rt
    M0       5      -0.294   32.5 0.966      0.965     239              249        152.9s        102.8s          50.0s  0.116    0.15       0.899             0.486
    M1       4       0.882   35.7  1.06      0.975     267              276        148.5s         94.7s          53.7s  0.136   0.172       0.558              1.67
    M2       4   -0.000808   33.5  1.06       0.96     242              260        141.3s         93.0s          48.3s  0.121   0.157       0.915             0.492
    M3       3       -1.84   37.6  2.28      0.784     276              312        305.3s        202.5s         102.7s     NA      NA          NA                NA
    M4       2     1.4e+03   17.1  19.4       0.97     223              226        323.5s        213.8s         109.7s     NA      NA          NA                NA

```{r}
dgtf_compare_params(list(
  M0 = fit,
  M1 = m1_fit,
  M2 = m2_fit,
  M3 = m3_fit,
  M4 = m4_fit
))
```

<dgtf_param_comparison: 5 parameters x 5 models, level = 95%>
 parameter model  mean median    sd  q_lo  q_hi
      par1    M0  1.06   1.06 0.161 0.735  1.37
      par1    M1 0.756  0.757 0.128 0.501     1
      par1    M2 0.482  0.484 0.058 0.367  0.59
      par2    M0 0.389  0.362 0.157 0.177 0.767
      par2    M1 0.363  0.343 0.123 0.184 0.666
       rho    M0  34.6   34.2  4.95  25.7  45.4
       rho    M1  26.6   26.4  3.43  20.5    34
       rho    M2  34.1   33.7  5.12  25.1    45
       rho    M3  25.4   25.1  3.97  18.7    34
      seas    M0  2.28   2.23 0.535   1.4   3.5
      seas    M1   2.2   2.12 0.601  1.24  3.56
      seas    M2  2.57   2.52 0.607  1.56  3.98
      seas    M3 0.404  0.362  0.22 0.097 0.929
      seas    M4 0.487  0.661  1.16 -2.24   2.3
         W    M0 0.012  0.012 0.001 0.009 0.015
         W    M2 0.013  0.012 0.002  0.01 0.016
         W    M3 0.004  0.004     0 0.003 0.005
         W    M4 0.173  0.168 0.037 0.114 0.257

```{r}
dgtf_compare(list(
  M0 = fc,
  M1 = m1_fc,
  M2 = m2_fc,
  M3 = m3_fc
  # M4 = m4_fc
))
```

<dgtf_comparison (dgtf_forecast): 4 rows, level = 95%>
 model horizon n_draws coverage    chi crps
    M0      10 500,000        1 0.0964   24
    M1      10 500,000        1  0.025 11.5
    M2      10 500,000      0.9  0.133 31.8
    M3      10 500,000        1  0.153 28.9

```{r}
dgtf_compare_plot(
  list(
    M0 = fc, 
    M1 = m1_fc, 
    M2 = m2_fc, 
    M3 = m3_fc, 
    True = ytest
  ),
  what = "yhat",
  time = length(ytrain) + seq_len(fc$horizon) # align x-axis with the forecast window
)

fits_list <- list(
  M0 = fit,
  M1 = m1_fit,
  M2 = m2_fit,
  M3 = m3_fit
)

fcs_list <- list(
  M0 = fc,
  M1 = m1_fc,
  M2 = m2_fc,
  M3 = m3_fc
)

plot_dgtf_y_compare(fits_list, fcs_list)
```

```{r}
plot_dgtf_lag(list(
  M0 = fit,
  M1 = m1_fit,
  M2 = m2_fit
), xlim = 21)
```

# 7. EpiEstim baseline

For comparison with a non-state-space estimator, the package's `epiestim_baseline()` wrapper runs [`EpiEstim::estimate_R()`][EpiEstim] with a sliding window and returns the `R_t` posterior bands in a format that `dgtf_compare_plot()` consumes directly. `EpiEstim` is in `Suggests`, so this section is conditional on having the package installed.

[EpiEstim]: https://cran.r-project.org/package=EpiEstim

```{r epiestim, eval = requireNamespace("EpiEstim", quietly = TRUE)}
# Serial-interval mean / sd implied by the Hawkes simulator's
# lognormal kernel `dgtf_hawkes(meanlog = 1.386, sigma2 = 0.323)`:
#   mean = exp(meanlog + sigma2 / 2)      ~= 4.7
#   sd   = mean * sqrt(exp(sigma2) - 1)   ~= 2.9
epi_baseline <- epiestim_baseline(
  y       = ytrain,
  mean_si = 4.7,
  sd_si   = 2.9,
  method  = "parametric_si",
  window  = 7
)
epi_baseline
```

<epiestim_baseline: method = parametric_si, window = 7, align = center, level = 0.95>
  SI mean = 4.7, SI sd = 2.9
  200 time points (7 NA, 193 estimated)

Last 6 rows:
           lower    median     upper
[195,] 0.7196854 0.7497249 0.7805888
[196,] 0.6779300 0.7079132 0.7387678
[197,] 0.6493553 0.6796007 0.7107710
[198,]        NA        NA        NA
[199,]        NA        NA        NA
[200,]        NA        NA        NA

```{r}
dgtf_compare_plot(
  list(
    M0       = fit,
    EpiEstim = epi_baseline,
    True     = Rt_train
  ),
  what = "Rt"
)
```