---
title: "Getting started with dgtf"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with dgtf}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6,
                      fig.height = 4)
```

```{r setup}
library(dgtf)
```

# 1. The DGTF model class

A **dynamic generalised transfer function (DGTF)** model has three pieces:

- an *observation* distribution (Poisson, negative-binomial, ...),
- a *link* function and *transfer / lag* structure that connects past
  observations and the latent state to the current intensity, and
- a *system* equation governing how the latent state evolves.

In `dgtf` each piece is its own component object, so you build a model
by snapping the parts together:

```{r build-model}
mod <- dgtf_model(
  obs         = obs_nbinom(),
  link        = link_identity(),
  sys         = sys_shift(),
  gain        = gain_softplus(),
  lag         = lag_lognormal(meanlog = 1.386262, sdlog = 0.3226017, window = 30),
  err         = err_gaussian(W = 0.01),
  seasonality = seas_period(7, init = c(1:7)),
  zi          = FALSE
)
mod
```

Most components also accept a string shortcut; the call above is
equivalent to `dgtf_model(obs = "nbinom", link = "identity",
sys = "shift", gain = "softplus", lag = lag_lognormal(...), ...)`.

For the canonical models discussed in the manuscript there are
one-line shortcuts:

```{r canonical}
dgtf_poisson_ar(p = 3)
dgtf_hawkes()
dgtf_koyck(kappa = 0.5)
dgtf_distributed_lag(r = 2, kappa = 0.5)
```

# 2. Simulate

```{r simulate}
sim <- dgtf_simulate_model(mod, ntime = 200, seed = 3269)
str(sim, max.level = 1)
plot(c(sim$y), type = "l", col = "grey50",
     xlab = "time", ylab = "y", main = "Simulated DGTF series")
```

# 3. Inference

The "what to infer" decision follows a single rule: anything you give a
prior is treated as unknown; anything you don't is fixed at the value
you supplied to `dgtf_model()`.

```{r infer}
prior <- dgtf_prior(
  seasonality = normal(1, 1),
  W           = inv_gamma(1, 1),
  rho         = inv_gamma(1, 1),
  lag         = list(par1 = normal(0, 1), par2 = inv_gamma(1, 1))
)

fit <- dgtf(
  y       = c(sim$y),
  model   = mod,
  prior   = prior,
  method  = "hva",
  control = vb_control(iter = 1000, n_sample = 1000, n_particle = 500)
)
Rt = log(exp(c(sim$psi)) + 1)
fit$error = dgtf_posterior_predictive(fit$fit, as_settings(mod), c(sim$y), 100, Rt)
```

# 4. Inspect

```{r summary}
summary(fit)
coef(fit)
```

```{r plot, fig.alt = "Posterior intensity vs simulated series"}
# In-sample posterior predictive check
ppc <- posterior_predict(fit, nrep = 100,
                         Rt = log1p(exp(c(sim$psi))))   # truth in a sim study

# Out-of-sample h-step-ahead forecast
fc  <- dgtf_forecast_fit(fit, h = 14, nrep = 200)

# With held-out truth for scoring
fc_eval <- dgtf_forecast_fit(fit, h = 14, nrep = 200,
                             ypred_true = sim$y[181:200])

# The standard predict() generic works too — delegates to the same engine
fc_via_predict <- predict(fit, horizon = 14, nrep = 200)
```

```{r diagnostics}
dgtf_diagnostics(fit)
```

# 5. Switch the inference method

The `model` and `prior` objects are reusable across methods — only the
control changes:

```{r switch-method, eval = FALSE}
fit_mcmc <- dgtf(sim$y, mod, prior,
                 method  = "mcmc",
                 control = mcmc_control(iter = 1000, warmup = 500))
```

# 6. Swap a component to change the model

The same data can be analysed under a different DGTF variant by
substituting one component:

```{r swap, eval = FALSE}
mod_geo  <- dgtf_hawkes(lag = lag_nbinom(r = 1, kappa = 0.5, window = 30))
mod_relu <- dgtf_hawkes(gain = gain_ramp())
```

A side-by-side comparison of all four canonical models on the same
dataset will appear in `vignette("four-models")`.
